DP Problems of Interest

### 2012-09-30 docs/roadmap/problems.tex
More problems.

% LaTeX Template for a short article
\documentclass[11pt]{article}
\usepackage{amssymb,amsmath,amsthm,hyperref,verbatim,pict2e,graphicx,marvosym,array,booktabs,nameref}
\hypersetup{colorlinks,citecolor=black,filecolor=black,linkcolor=black,urlcolor=black}
\usepackage[usenames]{xcolor} % color names ,dvipsnames,svgnames,table
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage[english]{babel}
\pagestyle{empty} % remove page numbers

@@

% Recurrence visualisation
\newcommand\Cd[3][0,-1]{\put(#2){\put(.5,.5){\circle*{.3}}\put(.5,.5){\linethickness{1.5pt}\vector(#1){#3}}}} % dependency [dx,dy]{x,y}{len}
\def\Cg#1{\put(#1){\color{lightgray}\put(0,0){\polygon*(0,0)(0,1)(1,1)(1,0)}}} % grayed cell (not to store
\def\Cz#1{\put(#1){\put(0,.35){\parbox{1\unitlength}{\centering\bf 0}}}} % zero-init cell
\def\Cm{\put(6.5,4.5){\circle*{.4}}\multiput(0,0)(1,0){9}{\line(0,1){8}}\multiput(0,0)(0,1){9}{\line(1,0){8}}} % matrix base

\title{DP problems of interest}
\author{Manohar Jonnalagedda, Thierry Coppey}
\date{}
\begin{document}
\maketitle
\pagestyle{headings}

@@

@@

% ------------------------------------------------------------------------------------------
% ------------------------------------------------------------------------------------------
% ------------------------------------------------------------------------------------------
{\color{red}
\item Matrices: $M_{n\times n}, B_{n \times n}$ \textit{<<first edge, last edge>>} upper triangular including main diagonal 
\item Alphabets:\ul
	\item Input: $\Sigma(S_{(i,j)})=\{0..m\}$ with $m=\max_S(i,j) \forall i,j$ determined at runtime\footnote{We need to scan/have stats about $S$ and that's where LMS plays a role}.
	\item Cost matrix: $\Sigma(M)=\{0..z\}$ with $z = m \cdot (n-2)$ (we add at most $n-2$ edges).
	\item Backtrack matrix: $\Sigma(B)=\{stop, 0..n\}$ (the index of the edge we add)
	\ule
\item Initialization: $M_{(i,i)}=0, B_{(i,i)}=stop \quad\forall i$
\item Recurrence: \[M_{(i,j)}=\left\{ \max_{i<k<j}M_{(i,k)}+M_{(k+1,j)} + S(i,k) \,\,\rule[-.75em]{.5pt}{2em}\,\,  k \right\} = B_{(i,j)} \]
	It is interesting to note that even in the sequential world, this problem is best solved 
	by filling the diagonals, ie. computing sub-solutions for all polygons on size $k$ before
	those of size $k+1$.
\item Backtracking: Starting at $B_{(0,n)}$ and let current point you backtrack drawing an edge from the current position to the backtracking position ??

\item Visualisation: in its traditional form the matrix is upper triangular with diagonal excluded.
\begin{center}\setlength{\unitlength}{.6cm}\begin{picture}(8,9)
	\put(-.7,6.5){\rotatebox{90}{First}}\put(-.4,6.4){\linethickness{1pt}\vector(0,-1){2}}
	\put(.2,8.2){Last}\put(1.5,8.4){\linethickness{1pt}\vector(1,0){2}}
	\Cg{0,6}
	\Cg{0,5} \Cg{1,5}
	\Cg{0,4} \Cg{1,4} \Cg{2,4}
	\Cg{0,3} \Cg{1,3} \Cg{2,3} \Cg{3,3}
	\Cg{0,2} \Cg{1,2} \Cg{2,2} \Cg{3,2} \Cg{4,2}
	\Cg{0,1} \Cg{1,1} \Cg{2,1} \Cg{3,1} \Cg{4,1} \Cg{5,1}
	\Cg{0,0} \Cg{1,0} \Cg{2,0} \Cg{3,0} \Cg{4,0} \Cg{5,0} \Cg{6,0}

	\Cz{0,7}\Cz{1,6}\Cz{2,5}\Cz{3,4}\Cz{4,3}\Cz{5,2}\Cz{6,1}\Cz{7,0}
	\Cd[0,1]{6,1}{2.8}\Cd[0,1]{6,2}{1.8}\Cd[0,1]{6,3}{0.8}
	\Cd[1,0]{3,4}{2.8}\Cd[1,0]{4,4}{1.8}\Cd[1,0]{5,4}{0.8}
	\put(3.5,4.5){\line(3,-1){3}}
	\put(4.5,4.5){\line(2,-2){2}}
	\put(5.5,4.5){\line(1,-3){1}}

	%\Cd[1,-1]{5,2}{0.8}

\Cm\end{picture}\end{center}

@@

% ----------------------------------------------
\newpage
\subsection{Matrix chain multiplication}\ol
\item Problem: find an optimal parenthesizing of the multiplication of $n$ matrices $A_i$. Each matrix $A_i$ is of dimension $r_i \times c_i$ \textit{<<r=rows, c=columns>>}. 
\item Matrices: $M_{n \times n}, B_{n \times n}$ \textit{(first, last matrix)}
\item Alphabets:\ul
	\item Input: matrix $A_i$ size is defined as pairs of integers $(r_i,c_i)$.
	\item Cost matrix: $\Sigma(M):=$ huge integer\footnote{Integer multiplication might blow up, use float or doubles and addition of logarithms instead.}.
	\item Backtrack matrix: $\Sigma(B)=\{stop\} \cup \{0..n\}$.
	\ule
\item Initialization:\ul
	\item Cost matrix: $M_{(i,i)}=0$.
	\item Backtrack matrix: $B_{(i,i)}=stop$.
	\ule

{\color{red}
\item Recurrence for the cost matrix:
	\[M_{(i,j)}=\min_{i\le k<j}\left\{\begin{array}{l|l}
		M_{(i,k)}+M_{(k,j)}+r_i \cdot r_k \cdot col_j & k
	\end{array}\right\}=B_{(i,j)} \]
\item Backtracking: XXX
\item Visualisation: XXX

\begin{center}\setlength{\unitlength}{.6cm}\begin{picture}(8,9)
	\put(-.7,6.5){\rotatebox{90}{First}}\put(-.4,6.4){\linethickness{1pt}\vector(0,-1){2}}
	\put(.2,8.2){Last}\put(1.5,8.4){\linethickness{1pt}\vector(1,0){2}}
	\Cg{0,6}
	\Cg{0,5} \Cg{1,5}
	\Cg{0,4} \Cg{1,4} \Cg{2,4}
	\Cg{0,3} \Cg{1,3} \Cg{2,3} \Cg{3,3}
	\Cg{0,2} \Cg{1,2} \Cg{2,2} \Cg{3,2} \Cg{4,2}
	\Cg{0,1} \Cg{1,1} \Cg{2,1} \Cg{3,1} \Cg{4,1} \Cg{5,1}
	\Cg{0,0} \Cg{1,0} \Cg{2,0} \Cg{3,0} \Cg{4,0} \Cg{5,0} \Cg{6,0}

	\Cz{0,7}\Cz{1,6}\Cz{2,5}\Cz{3,4}\Cz{4,3}\Cz{5,2}\Cz{6,1}\Cz{7,0}
	\Cd[0,1]{6,1}{2.8}\Cd[0,1]{6,2}{1.8}\Cd[0,1]{6,3}{0.8}
	\Cd[1,0]{3,4}{2.8}\Cd[1,0]{4,4}{1.8}\Cd[1,0]{5,4}{0.8}
	\put(3.5,4.5){\line(3,-1){3}}
	\put(4.5,4.5){\line(2,-2){2}}
	\put(5.5,4.5){\line(1,-3){1}}

	%\Cd[1,-1]{5,2}{0.8}

\Cm\end{picture}\end{center}

\item Optimizations: XXX
}
\ole

% ----------------------------------------------
\newpage
\subsection{Nussinov algorithm}\ol
\item Problem: folding a RNA string $S$ over itself.
{\color{red}
\item Matrices: XXX
\item Alphabets:\ul
	\item Input: $\Sigma(S)=\{A,C,G,U\}$.
	\item Cost matrix: $\Sigma(M)=\{0..n\}, n={\rm length}(S)/2$
	\item Backtrack matrix: $\Sigma(B)=\{stop,W,S,SW, 1..n\}$
	\ule

\item Initialization: \ul
	\item Cost matrix: $\left\{\begin{array}{l} M_{(i,i)}=0 \\ M_{(i,i-1)}=0 \end{array}\right. \forall i \in 1..{\rm length}(S)$
	\item Backtrack matrix: $\left\{\begin{array}{l} B_{(i,i)}=stop \\ B_{(i,i-1)}=stop \end{array}\right.  \forall i \in 1..{\rm length}(S)$
	\ule

\item Recurrences:
	\[M_{(i,j)}=\max\left\{\begin{array}{l|l}
		M_{(i+1,j-1)}+\omega(i,j) & SW\\
		M_{(i+1,j)} & S\\
		M_{(i,j-1)} & W\\
		\max_{i<k<j}M_{(i,k)}+M_{(k+1,j)} & k
	\end{array}\right\} = B_{(i,j)} \]
	With $\omega(i,j)=1$ if $i,j$ are complementary. 0 otherwise.
\item Backtracking: XXX
\item Visualisation: XXX
\item Optimizations: XXX
}
\ole

% ------------------------------------------------------------------------------------------------
\newpage
\subsection{Zuker folding}\ol
\item Problem: folding a RNA string $S$ over itself.
{\color{red}
\item Matrices: XXX

\item Alphabets:\ul
\item Input: $\Sigma(S)=\{A,C,G,U\}$.
\item Cost matrix: $\Sigma(W)=\Sigma(V)=\Sigma(F)=\{0..n\},n=$
\item Backtrack matrix: $\Sigma(B)=\{\}$
\ule

\item Initialization:\ul
	\item Cost matrices:\ul
		\item XXX: $W$,$V$
		\item $F_{(0)}=0$
	\ule
	\item Backtrack matrix: $B$
	\ule

>>>>

\documentclass[11pt]{article}
\usepackage{amssymb,amsmath,amsthm,hyperref,verbatim,pict2e,graphicx,marvosym,array,booktabs,nameref}
\hypersetup{colorlinks,citecolor=black,filecolor=black,linkcolor=black,urlcolor=black}
\usepackage[usenames]{xcolor} % color names ,dvipsnames,svgnames,table
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage[english]{babel}
\pagestyle{empty} % remove page numbers

@@

% Recurrence visualisation helpers
\newcommand\Cd[3][0,-1]{\put(#2){\put(.5,.5){\circle*{.3}}\put(.5,.5){\linethickness{1.5pt}\vector(#1){#3}}}} % dependency [dx,dy]{x,y}{len}
\def\Cg#1{\put(#1){\color{lightgray}\put(0,0){\polygon*(0,0)(0,1)(1,1)(1,0)}}} % grayed cell (not to store
\def\Cz#1{\put(#1){\put(0,.35){\parbox{1\unitlength}{\centering\bf 0}}}} % zero-init cell
\def\Cm{\put(6.5,4.5){\circle*{.4}}\multiput(0,0)(1,0){9}{\line(0,1){8}}\multiput(0,0)(0,1){9}{\line(1,0){8}}} % matrix base
\def\Cfl#1{#1{0,6}#1{0,5}#1{1,5}#1{0,4}#1{1,4}#1{2,4}#1{0,3}#1{1,3}#1{2,3}#1{3,3}#1{0,2}#1{1,2}#1{2,2}#1{3,2}#1{4,2}
	#1{0,1}#1{1,1}#1{2,1}#1{3,1}#1{4,1}#1{5,1}#1{0,0}#1{1,0}#1{2,0}#1{3,0}#1{4,0}#1{5,0}#1{6,0}} % triangular lower (function)
\def\Cfd#1{#1{0,7}#1{1,6}#1{2,5}#1{3,4}#1{4,3}#1{5,2}#1{6,1}#1{7,0}} % main diagonal

\title{DP Problems of Interest}
\author{Manohar Jonnalagedda, Thierry Coppey}
\date{}
\begin{document}
\maketitle
\pagestyle{headings}

@@

% ----------------------------------------------
\newpage
\subsection{Smith-Waterman with arbitrary gap cost (3D)}\ol

\item Problem: matching two strings $S$, $T$ with $|S|=m, |T|=n$ with an arbitrary gap function. Without loss of generality, let $m\ge n$\footnote{Otherwise if $|T|>|N|$ we only need to swap both the inputs and backtracking pairs.}. Example penalty function could be\footnote{Intuition: long gaps penalize less, at some point, one large gap is better than matching and smaller gaps.} $g(x)=m-x$.

\item Matrices: $M_{m \times n \times m}, B_{m \times n \times m}$
\item Alphabets:\ul
	\item Input: $\Sigma(S)=\Sigma(T)=\{a,c,g,t\}$.
	\item Cost matrix: $\Sigma(M) = [0..z], z=\max({\rm cost(\_)}) \cdot \min(m,n)$
	\item Backtrack matrix: $\Sigma(B)=\{stop,W,N,NW\}$
	\ule

{\color{red}
\item Initialization: XXX: we build this, because we go 3D.
\item Recurrences:
\item Backtracking:
\item Visualisation:
\item Optimizations:
}
\ole

@@

{\color{red}
\item Matrices: $M_{n\times n}, B_{n \times n}$ \textit{<<first edge, last edge>>} upper triangular including main diagonal 
\item Alphabets:\ul
	\item Input: $\Sigma(S_{(i,j)})=\{0..m\}$ with $m=\max_S(i,j) \forall i,j$ determined at runtime\footnote{We need to scan/have stats about $S$ and that's where LMS plays a role}.
	\item Cost matrix: $\Sigma(M)=\{0..z\}$ with $z = m \cdot (n-2)$ (we add at most $n-2$ edges).
	\item Backtrack matrix: $\Sigma(B)=\{stop, 0..n\}$ (the index of the edge we add)
	\ule
\item Initialization: $M_{(i,i)}=0, B_{(i,i)}=stop \quad\forall i$
\item Recurrence: \[M_{(i,j)}=\left\{ \max_{i<k<j}M_{(i,k)}+M_{(k+1,j)} + S(i,k) \,\,\rule[-.75em]{.5pt}{2em}\,\,  k \right\} = B_{(i,j)} \]
	It is interesting to note that even in the sequential world, this problem is best solved 
	by filling the diagonals, ie. computing sub-solutions for all polygons on size $k$ before
	those of size $k+1$.
\item Backtracking: Starting at $B_{(0,n)}$ and let current point you backtrack drawing an edge from the current position to the backtracking position ??

\item Visualisation: in its traditional form the matrix is upper triangular with diagonal excluded.
\begin{center}\setlength{\unitlength}{.6cm}\begin{picture}(8,9)
	\put(-.7,6.5){\rotatebox{90}{First}}\put(-.4,6.4){\linethickness{1pt}\vector(0,-1){2}}
	\put(.2,8.2){Last}\put(1.5,8.4){\linethickness{1pt}\vector(1,0){2}}
	\Cfl{\Cg}\Cfd{\Cz}
	\Cd[0,1]{6,1}{2.8}\Cd[0,1]{6,2}{1.8}\Cd[0,1]{6,3}{0.8}
	\Cd[1,0]{3,4}{2.8}\Cd[1,0]{4,4}{1.8}\Cd[1,0]{5,4}{0.8}
	\put(3.5,4.5){\line(3,-1){3}}
	\put(4.5,4.5){\line(2,-2){2}}
	\put(5.5,4.5){\line(1,-3){1}}
	%\Cd[1,-1]{5,2}{0.8}
\Cm\end{picture}\end{center}

@@

% ----------------------------------------------
\newpage
\subsection{Matrix chain multiplication}\ol
\item Problem: find an optimal parenthesizing of the multiplication of $n$ matrices $A_i$. Each matrix $A_i$ is of dimension $r_i \times c_i$ and $c_i=r_{i+1} \forall i$. \textit{<<r=rows, c=columns>>}
\item Matrices: $M_{n \times n}, B_{n \times n}$ \textit{(first, last matrix)}
\item Alphabets:\ul
	\item Input: matrix $A_i$ size is defined as pairs of integers $(r_i,c_i)$.
	\item Cost matrix: $\Sigma(M):=$ huge integer\footnote{Integer multiplication might blow up, use float or doubles and addition of logarithms instead.}.
	\item Backtrack matrix: $\Sigma(B)=\{stop\} \cup \{0..n\}$.
	\ule
\item Initialization:\ul
	\item Cost matrix: $M_{(i,i)}=0$.
	\item Backtrack matrix: $B_{(i,i)}=stop$.
	\ule
\item Recurrence: $c_k=r_{k+1}$
	\[M_{(i,j)}=\min_{i\le k<j}\left\{\begin{array}{l|l}
		M_{(i,k)}+M_{(k+1,j)}+r_i \cdot c_k \cdot c_j & k
	\end{array}\right\}=B_{(i,j)} \]
\item Backtracking: Start at $B_{(1,n)}$. Use the following recursive function for parenthesizing
	\[{\rm BT}(B_{(i,j)}=k) \mapsto \left\{\begin{array}{ll} A_i & \text{if } k=0 \lor k=j \\
		\Big( {\rm BT}(B_{(i,k)}) \Big) \cdot \Big( {\rm BT}(B_{(k+1,j)}) \Big) & \text{otherwise} \end{array}\right.\]

\item Visualisation:
	\begin{center}\setlength{\unitlength}{.6cm}\begin{picture}(8,9)
		\put(-.7,6.5){\rotatebox{90}{First}}\put(-.4,6.4){\linethickness{1pt}\vector(0,-1){2}}
		\put(.2,8.2){Last}\put(1.5,8.4){\linethickness{1pt}\vector(1,0){2}}
		\Cfl{\Cg}\Cfd{\Cz}
		\Cd[0,1]{6,1}{2.8}\Cd[0,1]{6,2}{1.8}\Cd[0,1]{6,3}{0.8}
		\Cd[1,0]{3,4}{2.8}\Cd[1,0]{4,4}{1.8}\Cd[1,0]{5,4}{0.8}
		\put(3.5,4.5){\line(3,-1){3}}\put(4.5,4.5){\line(2,-2){2}}\put(5.5,4.5){\line(1,-3){1}}
	\Cm\end{picture}\end{center}

\item Optimizations:\ul
	\item {\color{red}We need to swap vertically the matrix to have a normalized progression towards bottom right. To do that, we need to map all indices $i \mapsto n-1-i$, but since we want to store the matrix sparsely, we might want to transform it into a lower triangular matrix as we can provide a size-independent mapping of element indices.
	XXX: what's the best trade off ? progress towards bottom right VS map indices more efficiently ?}
	\ule
\ole

% ----------------------------------------------
\newpage
\subsection{Nussinov algorithm}\ol
\item Problem: folding a RNA string $S$ over itself $\left\lfloor |S| / 2 \right\rfloor = n$.
\item Matrices: $M_{n\times n}, B_{n \times n}$
\item Alphabets:\ul
	\item Input: $\Sigma(S)=\{A,C,G,U\}$.
	\item Cost matrix: $\Sigma(M)=\{0..n\}$
	\item Backtrack matrix: $\Sigma(B)=\{stop,W,S,SW, 1..n\}$
	\ule
\item Initialization: \ul
	\item Cost matrix: $ M_{(i,i)}=M_{(i,i-1)}=0$
	\item Backtrack matrix: $B_{(i,i)}=B_{(i,i-1)}=stop$
	\ule

{\color{red}
\item Recurrences (XXX: isn't it $\max_{i\le k < j}$ ?):
	\[M_{(i,j)}=\max\left\{\begin{array}{l|l}
		M_{(i+1,j-1)}+\omega(i,j) & SW\\
		M_{(i+1,j)} & S\\
		M_{(i,j-1)} & W\\
		\max_{i<k<j}M_{(i,k)}+M_{(k+1,j)} & k
	\end{array}\right\} = B_{(i,j)} \]
	With $\omega(i,j)=1$ if $i,j$ are complementary. 0 otherwise.
\item Backtracking: XXX
\item Visualisation: XXX
\item Optimizations: XXX
}
\ole

% ------------------------------------------------------------------------------------------------
\newpage
\subsection{Zuker folding}\ol
\item Problem: folding a RNA string $S$ over itself $\left\lfloor |S| / 2 \right\rfloor = n$.
\item Matrices: $M_{n\times n}, V_{n\times n}, B_{n \times n}, {\color{red} F???}$
\item Alphabets:\ul
	\item Input: $\Sigma(S)=\{A,C,G,U\}$.
	\item Cost matrix: $\Sigma(W)=\Sigma(V)=\Sigma(F)=\{0..n\}$
	\item Backtrack matrix: $\Sigma(B)=\{stop, {\color{red} S,W, XXX here} \}$
	\ule
{\color{red}
\item Initialization:\ul
	\item Cost matrices:\ul
		\item XXX: $W$,$V$
		\item $F_{(0)}=0$
	\ule
	\item Backtrack matrix: $B$
	\ule
